---
title: "Functions and Data Handling in R"
output:
  html_notebook:
    toc: true
    toc_depth: 3
---

## Learning Objectives

After completing this lab, students will be able to:

- Create and use variables in R
- Write simple user-defined functions
- Read data from CSV files
- Explore and inspect datasets
- Create new variables from existing data
- Save modified datasets to files

## 1. Creating Variables

Variables store data values.

**Numeric Variables**

```{r}
age <- 22 
cgpa <- 3.85 
age 
cgpa
```
**Character Variables**

```{r}
name <- "Rahim" 
department <- "CSE" 
name 
department
```

**Logical Variables**

```{r}
passed <- TRUE
passed
```

## 2. Working with Vectors
A vector stores multiple values of the same type.

```{r}
marks <- c(85, 78, 90, 88, 76)
marks
```
Access elements:
```{r}
marks[1]
marks[3]
```

Basic operations:

```{r}
mean(marks)
max(marks)
min(marks)
sum(marks)
```

## 3. Writing Simple Functions
Functions help reuse code.

General syntax:

function_name <- function(parameters){

    statements

    return(value)

}

**Example 1: Addition Function**
```{r}
add_numbers <- function(a, b){ 
  result <- a + b 
  return(result) 
} 

add_numbers(10, 20)
```

**Example 2: Area of Rectangle**
```{r}
rectangle_area <- function(length, width){
  area <- length * width
  return(area)
}

rectangle_area(10, 5)
```

**Example 3: Grade Function**
```{r}
calculate_grade <- function(score){

  if(score >= 80){
    return("A")
  }

  else if(score >= 70){
    return("B")
  }

  else if(score >= 60){
    return("C")
  }

  else{
    return("F")
  }

}

calculate_grade(85)
```

## 4. Built-in Dataset Exploration

R provides several built-in datasets.

Load the iris dataset:
```{r}
data(iris)
```

Display first rows:
```{r}
head(iris)
```

Display last rows:
```{r}
tail(iris)
```

## 5. Basic Dataset Inspection
Structure of Dataset
```{r}
str(iris)
```

Shows:

- Variable names
- Data types
- Sample values

**Dataset Dimensions**
```{r}
dim(iris)
```

Output:

`150 rows and 5 columns`

**Column Names**
```{r}
names(iris)
```


**Summary Statistics**
```{r}
summary(iris)
```

**View Dataset**
```{r}
View(iris)
```

## 6. Reading Data from Files

**Reading CSV Files**

students.csv

Example:

| ID | Name   | CGPA |
|----|--------|------|
| 1  | Rahim  | 3.70 |
| 2  | Karim  | 3.50 |
| 3  | Nabila | 3.90 |

**Read CSV File**
```{r}
# Change the path for your file
students <- read.csv("D:\\IDS\\Lab\\new\\students.csv")
students
```
Alternative (Recommended)
```{r}
library(readr)
students <- read_csv("D:\\IDS\\Lab\\new\\students.csv", show_col_types = FALSE)
students
```

Display data:
```{r}
head(students)
```

**Check Structure**
```{r}
str(students)
```

**Reading Text Files**

Example: students.txt

ID Name CGPA

1 Rahim 3.70

2 Karim 3.50

3 Nabila 3.90

```{r}
#Read Text File
students <- read.table(
    "D:\\IDS\\Lab\\new\\students.txt",
    header = TRUE
)

head(students)
```

**Read Tab-Separated Text File**

Example:

ID	Name	CGPA

1	Rahim	3.70

2	Karim	3.50

3	Nabila	3.90

```{r}
students <- read.delim(
    "students.txt"
)

head(students)
```

**Reading Excel Files (.xlsx)**

```{r}
#Install package:
install.packages("readxl")

#Load package:
library(readxl)

#Read Excel file:

students <- read_excel(
    "students.xlsx"
)

head(students)
```

Read Specific Worksheet
```{r}
students <- read_excel(
  # change the path
    "students.xlsx",
    sheet = "Sheet1"
)
```

**Reading Data from URLs**

Many datasets are available online.

Example:
```{r}

url <- "https://raw.githubusercontent.com/mwaskom/seaborn-data/master/iris.csv"

iris_data <- read.csv(url)

head(iris_data)
```

**Reading Data from GitHub**

GitHub hosts many publicly available datasets.
```{r}
github_url <- "https://raw.githubusercontent.com/cs109/2014_data/master/countries.csv"

countries <- read.csv(
    github_url
)

head(countries)
```

**Reading Data from Google Drive**

Read a Public CSV Directly from Google Drive

Suppose the Google Drive sharing link is:

https://drive.google.com/file/d/FILE_ID/view?usp=sharing

Example:

https://drive.google.com/file/d/1ABCDEF123456789/view?usp=sharin

<span style="color:red;">
The file must be shared publicly or accessible to your account.
</span>

The File ID is: 
1ABCDEF123456789

Create a direct download URL:
```{r}
file_id <- "1ABCDEF123456789"

url <- paste0(
  "https://drive.google.com/uc?export=download&id=",
  file_id
)

#if the file is CSV
students <- read.csv(url) 

head(students)
```

## 7. Creating New Variables

Suppose the dataset contains:

students

Create a scholarship variable:
```{r}
students$Scholarship <- students$CGPA >= 3.75
```

Display:
```{r}
students
```

**Create Grade Category**
```{r}
students$Grade <- ifelse(
    students$CGPA >= 3.75,
    "Excellent",
    "Good"
)
```

Display:
```{r}
students
```



## 8. Writing Data into Files

Save dataset as CSV.
```{r}
write.csv(
    students,
    "students_updated.csv",
    row.names = FALSE
)
```

A new file named:students_updated.csv will be created.

Since this is only the **second lab** of an undergraduate Introduction to Data Science course, the exercises should gradually move from **variables → functions → reading data → data inspection → creating variables → saving data**.

I would organize them into three levels: Basic, Intermediate, and Challenge.


## Lab Exercises

### Exercise 1: Creating Variables

Create the following variables:

* Student Name
* Student ID
* Age
* CGPA

Display all variables.

Expected output:

```text
Name: Rahim
ID: 221-15-1234
Age: 22
CGPA: 3.85
```

### Exercise 2: Working with Vectors

Create a vector containing the following quiz marks:

```text
80, 75, 90, 85, 95
```

Find:

* Total marks
* Average marks
* Highest mark
* Lowest mark

### Exercise 3: Creating Functions

Write a function that calculates:

```text
Area = Length × Width
```

Test the function using:

```text
Length = 12
Width = 8
```


### Exercise 4: Temperature Conversion

Write a function that converts Celsius to Fahrenheit.

Formula:

```text
F = (9/5 × C) + 32
```

Test using:

```text
C = 25
```

### Exercise 5: Exploring the mtcars Dataset

Load the built-in dataset:

```r
data(mtcars)
```

Perform the following tasks:

1. Display the first 10 rows.
2. Display the last 10 rows.
3. Determine the number of rows and columns.
4. Display column names.
5. Display summary statistics.

### Exercise 6: Dataset Inspection

Using the mtcars dataset:

Display:

```r
head(mtcars)
tail(mtcars)
str(mtcars)
summary(mtcars)
```

Briefly explain what each function does.

### Exercise 7: Reading a CSV File

Suppose the file:

```text
students.csv
```

contains:

| ID | Name   | CGPA |
| -- | ------ | ---- |
| 1  | Rahim  | 3.70 |
| 2  | Karim  | 3.50 |
| 3  | Nabila | 3.90 |

Read the file and:

1. Display the first rows.
2. Display its structure.
3. Display summary statistics.

### Exercise 8: Reading an Excel File

Read:

```text
students.xlsx
```

Display:

```r
head(data)
summary(data)
```

### Exercise 9: Reading Data from an Online Source

Load the Iris dataset directly from:

```r
url <- "https://raw.githubusercontent.com/mwaskom/seaborn-data/master/iris.csv"
```

Display:

* First 5 rows
* Dataset dimensions
* Summary statistics


### Exercise 10: Scholarship Eligibility

Using the following dataset:

| ID | Name   | CGPA |
| -- | ------ | ---- |
| 1  | Rahim  | 3.80 |
| 2  | Karim  | 3.40 |
| 3  | Nabila | 3.95 |

Create a new variable:

```text
Scholarship
```

Rule:

```text
CGPA ≥ 3.75 → TRUE
Otherwise → FALSE
```

### Exercise 11: Performance Category

Create a variable:

```text
Performance
```

Rule:

```text
CGPA ≥ 3.75      → Excellent
3.50–3.74        → Good
Below 3.50       → Average
```

Display the updated dataset.

### Exercise 12: Save a Modified Dataset

After creating the Scholarship and Performance variables:

Save the dataset as:

```text
students_updated.csv
```

Verify that the file is created successfully.


## Mini Project

Create a dataset for at least 10 students containing:

* ID
* Name
* Quiz Marks
* Assignment Marks
* Midterm Marks

Perform the following:

1. Read the dataset.
2. Inspect the dataset.
3. Create a Total Marks variable.
4. Create a Grade variable.
5. Display summary statistics.
6. Save the updated dataset as:

```text
final_results.csv
```

