---
title: "Data Preprocessing"
author: "Roll: Clas ROll,ID: AB-XYZXY-CD, Name: Your Full Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
  pdf_document:
    toc: true
---

# Learning Objectives

After completing this lab, students will be able to:

- Identify and handle missing values
- Detect and remove duplicate records
- Identify inconsistent data
- Detect and handle outliers
- Perform data wrangling operations
- Encode categorical variables
- Normalize and standardize numerical variables
- Create new features
- Perform feature selection
- Apply dimensionality reduction
- Perform numerosity reduction through sampling

# 1. Install and Load Packages

```{r}
required_packages <- c(
  "tidyverse",
  "janitor",
  "caret",
  "FactoMineR",
  "factoextra"
)

for(pkg in required_packages) {

  if(!require(pkg, character.only = TRUE)) {

    cat("Installing", pkg, "...\n")

    install.packages(pkg)

    library(pkg, character.only = TRUE)

  } else {

    cat(pkg, "is already installed.\n")

  }

}
```

# 2. Load Dataset
```{r}
titanic <- read.csv("titanic.csv")
head(titanic)
str(titanic)
```
# Part I: Data Cleaning
## 3. Missing Values
**Identify Missing Values**

```{r}
colSums(is.na(titanic))
```

**Percentage of Missing Values**

```{r}
round(
  colSums(is.na(titanic)) /
  nrow(titanic) * 100,
  2
)
```
**Remove Missing Records**

```{r}
titanic_clean <- na.omit(titanic)
```

**Impute Missing Age**

```{r}
titanic$Age[
  is.na(titanic$Age)
] <- mean(
  titanic$Age,
  na.rm = TRUE
)
```

## 4. Duplicate Records
**Check Duplicates**
```{r}
sum(
  duplicated(titanic)
)
```

**Remove Duplicates**
```{r}
titanic <- distinct(titanic)
```

## 5. Inconsistent Data
**Example**
```{r}
unique(
  titanic$Sex
)
```

male

Male

MALE

would be inconsistent.

**Standardize Values**
```{r}
titanic$Sex <- tolower(
  titanic$Sex
)
```

## 6. Outlier Detection
**Boxplot**

```{r}
boxplot(
  titanic$Fare,
  main = "Fare Distribution"
)
```
**IQR Method**

```{r}
Q1 <- quantile(
  titanic$Fare,
  0.25
)

Q3 <- quantile(
  titanic$Fare,
  0.75
)

IQR_value <- IQR(
  titanic$Fare
)

lower <- Q1 - 1.5 * IQR_value
upper <- Q3 + 1.5 * IQR_value

outliers <- titanic$Fare[
  titanic$Fare < lower |
  titanic$Fare > upper
]

outliers
```
# Part II: Data Transformation
## 7. Data Wrangling
**Tidy Data Check**

```{r}
head(titanic)
```

Discuss:

Variables in columns

Observations in rows

## 8. Filtering
Passengers older than 30
```{r}
titanic %>%
  filter(Age > 30)
```

## 9. Sorting
Highest Fare

```{r}
titanic %>%
  arrange(desc(Fare))
```

## 10. Joins and Merges
Create a small example:
```{r}
passengers <- data.frame(
  PassengerId = c(1,2,3),
  Name = c(
    "John",
    "Sara",
    "Tom"
  )
)

tickets <- data.frame(
  PassengerId = c(1,2,3),
  TicketClass = c(
    "First",
    "Second",
    "Third"
  )
)

merge(
  passengers,
  tickets,
  by = "PassengerId"
)
```

## 11. Reshaping Data
```{r}
scores <- data.frame(
  Student = c(
    "Rahim",
    "Karim"
  ),
  Quiz1 = c(80,70),
  Quiz2 = c(85,75)
)

scores
```

Wide → Long

```{r}
pivot_longer(
  scores,
  cols = c(
    Quiz1,
    Quiz2
  ),
  names_to = "Quiz",
  values_to = "Marks"
)
```

## 12. Encoding Categorical Variables
**Convert to Factors**
```{r}
titanic$Sex <- as.factor(
  titanic$Sex
)

titanic$Embarked <- as.factor(
  titanic$Embarked
)
```
**Numeric Encoding**

```{r}
titanic$Sex_Code <-
  as.numeric(
    titanic$Sex
  )
```

## 13. Normalization
**Min-Max Scaling**
```{r}
titanic$Age_Norm <-
(
  titanic$Age -
  min(titanic$Age)
) /
(
  max(titanic$Age) -
  min(titanic$Age)
)
```

## 14. Standardization
```{r}
titanic$Age_Std <-
  scale(
    titanic$Age
  )
```

## 15. Feature Engineering
Create Family Size
```{r}
titanic$FamilySize <-
  titanic$SibSp +
  titanic$Parch + 1
```

Create Child Indicator

```{r}
titanic$IsChild <-
  ifelse(
    titanic$Age < 18,
    1,
    0
  )
```

# Part III: Data Reduction
## 16. Feature Selection
Select Important Variables
```{r}
selected_data <-
  titanic %>%
  select(
    Survived,
    Age,
    Fare,
    Sex,
    Pclass
  )
```
## 17. Dimensionality Reduction (PCA)
Create Numerical Dataset
```{r}
numeric_data <-
  titanic %>%
  select_if(is.numeric)
```
Run PCA

```{r}
pca_result <-
  prcomp(
    numeric_data,
    scale. = TRUE
  )
```

Summary

```{r}
summary(
  pca_result
)
```
PCA Plot
```{r}
fviz_pca_ind(
  pca_result
)
```

## 18. Numerosity Reduction
**Random Sampling**

Original Size

```{r}
nrow(titanic)
```
Take 20% Sample

```{r}
sample_data <-
  sample_n(
    titanic,
    size =
      0.2 *
      nrow(titanic)
  )
```

Sample Size
```{r}
nrow(sample_data)
```

